---
title: "Introduction to the Australian System of National Accounts"
subtitle: "Module 1: Gross Domestic Product (Expenditure)"
author: "David Hansell"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(kableExtra)
library(readabs)
library(tidytable)
library(ggplot2)
library(tidyverse)
library(scales, lubridate)
library(data.table)
library(plotly)

gov_c <- read_abs(series_id = "A2420654T") %>% 
  select.(value, date) %>% 
  rename.(`General Government Consumption` = value)
hhold_c <- read_abs(series_id = "A2420655V") %>% 
  select.(value, date) %>% 
  rename.(`Household Consumption` = value)
public_i <- read_abs(series_id = "A2420683C") %>% 
  select.(value, date) %>% 
  rename.(`Public Investment` = value)
private_i <- read_abs(series_id = "A2420674A") %>% 
  select.(value, date) %>% 
  rename.(`Private Investment` = value)
inventories <- read_abs(series_id = "A2420689T") %>% 
  select.(value, date) %>% 
  mutate.(value = ifelse(is.na(value), 0, value)) %>% 
  rename.(`Inventories` = value)
exports <- read_abs(series_id = "A2420691C") %>% 
  select.(value, date) %>% 
  rename.(Exports = value)
imports <- read_abs(series_id = "A2420692F") %>% 
  select.(value, date) %>% 
  rename.(Imports = value)
stat_disc <- read_abs(series_id = "A2420693J") %>% 
  select.(value, date) %>% 
  rename.(`Statisitical Discrepancy` = value)

gdp_e <- read_abs(series_id = "A2420912W") %>% 
  select.(value, date) %>% 
  rename.(`GDP (E)` = value)

gdp_parts <- gov_c %>% 
  full_join.(hhold_c) %>% 
  full_join.(public_i) %>% 
  full_join.(private_i) %>% 
  full_join.(inventories) %>% 
  full_join.(exports) %>% 
  full_join.(imports) %>% 
  full_join.(stat_disc) %>% 
  full_join.(gdp_e) %>% 
  mutate.(`GDP (E) - Sum of Parts` = `General Government Consumption` + 
            `Household Consumption` + 
            `Public Investment` + 
            `Private Investment` + 
            Inventories + Exports - Imports + `Statisitical Discrepancy`, 
          `Net Exports` = Exports - Imports) %>% 
  mutate.(dif = `GDP (E)` - `GDP (E) - Sum of Parts`)

```
## Aims of the module

1. Understand the expenditure approach of GDP 
2. How to map these concepts to ABS national accounts time series


Note: there is a corresponding module that demonstrates how to download ABS data directly in R-Studio and make the charts presented here.

## Overview

- GDP = Gross Domestic Product
  + **Gross** = includes depreciation of capital
  + **Domestic** = confined to a production boundary
  + **Product** = what is produced
- System of National Accounts (SNA) defines GDP three ways
  + **Expenditure**: Consumption + Investment + Govt Purchases + Exports – Imports 
  + **Income**: Compensation of Employees (CoE) + Gross Operating Surplus (GOS)
  + **Production**: Output less intermediate Expenses (value-added). 
- In other words, we spend our income on what we produce.

## Overview

This module focusses on the expenditure approach or GDP(E). 

We divide the module into two parts:  

1. Concepts  
2. Measurement.
  
# Concepts

## GDP (E) - Concepts

Expenditure approach to GDP is based on the framework: 
$$
\begin{aligned}
GDP = \text{Consumption}   \\
+ \text{ Investment} \\
+ \text{ Govt purchases of goods and services} \\
+ \text{ Exports} \\
- \text{ Imports} 
\end{aligned}
$$
This is often abbreviated as $GDP = C + I + G + X - M$

## GDP (E) - Concepts

- **Consumption** means the good or service is used within the period of observation. Thus expenditure is final.
- **Investment** implies the good or service (incl. capitalized wages) is funneled into an asset that lasts for a while so it makes things/provides services in more than one period.
  - In other words, investment is a flow that creates a stock (capital).
- **Governments** employ people, tax and spend, but households receive income to consume so we don’t count the wages of govt employees as govt expenditure (unless they’re transformed into capital, see above). 
  - We only include government purchases and then separate these into final consumption (eg. stationery) or investment (eg. roads).

## GDP (E) - Concepts

- **Production boundary**: The geographic area of concern. In this case, Australia and associated territories. 
  - Note: the ABS national accounts also include state accounts, which are just subsets of the broader set.
- **Exports** are goods or services within the production boundary
- **Imports** are goods or services produced outside the production boundary.  

# Measurement

## Measurement - Overview

- ABS publishes [annual (5204)](https://www.abs.gov.au/statistics/economy/national-accounts/australian-system-national-accounts/latest-release) and [quarterly (5206)](https://www.abs.gov.au/statistics/economy/national-accounts/australian-national-accounts-national-income-expenditure-and-product/latest-release) national accounts. 
- Quarterly estimates are often carried forward by projecting the movements in correlated series and are subject to more revision than annual estimates. 
- Annual estimates (apart from the most recent year) are balanced via the supply-use tables. 
  - Main point: Once the level of GDP is finalized via balancing the statistical discrepancy item is zero. Supply-use tables start in 1994 so before this the statistical discrepancy is not zero.
- Hence here we focus on the annual estimates found in table 2 of catalogue 5204.

## Measurement - Overview

- Table 2 contains chain volume estimates of expenditure. 
  - Current price estimates for expenditure are published in the quarterly accounts (5206)
- Table 2 contains 174 series. Many of these are subtotals so we can’t add up the rows to get the total. 
- More detailed breakdowns of these series are found in other tables of the 5204 publication. 
  - For example, table 8 contains a detailed breakdown of household final consumption expenditure.
- Fortunately many of these are not necessary to get a high-level of understanding. In fact we only need 8 columns and not all 8 are of equal importance. 

## Measurement - Consumption

- The two main components of consumption expenditure are
  - **Household Consumption**:
    - Series ID = A2420655V
    - Official title: Households; Final Consumption Expenditure
  - **General Government Consumption**: 
    - Series ID = A2420654T
    - Official title: General Government; Final Consumption Expenditure
    - Encompasses all levels of Australian government.
- Note: The G in the formula on slide 3 is separated into consumption and investment in the national accounts.

## Measurement - Investment

- **Investment** in the SNA is referred to as *gross fixed capital formation*. The two main series are
  - **Private Investment**: 
    - Series ID = A2420674A
    - Official title: Private; Gross Fixed Capital Formation
  - **Public Investment**: 
    - Series ID = A2420683C
    - Official title: Public; Gross Fixed Capital Formation
    
## Measurement - Exports and Imports

- **Exports** count towards GDP, while **imports** are subtracted because they weren’t produced within the Australian production boundary. The two main series are
  - **Exports**: 
    - Series ID = A2420691C
    - Official title: Export of Goods and Services
  - **Imports**: 
    - Series ID = A2420692F
    - Official title: Imports of Goods and Services
    
## Measurement - Inventories and Discrepancy

- **Inventories** can be negative or positive as goods or services produced in previous time periods can be run down (ie, negative) or added to in the current period (ie, positive)
  - Series ID = A2420689T
  - Official title: Change in inventories
- **Statistical discrepancy** refers to the difference in this account with the others. It disappears after the ABS conducts supply-use balancing, ie, it apportions the difference to either accounts and hence all measures of GDP become equal.
  - Series ID = A2420693J
  - Official title: Statistical Discrepancy (E)

## Measurement - Recap

$$
\begin{aligned}
GDP (E) = \text{Household Consumption}   \\
+ \text{ General Government Consumption} \\
+ \text{ Private Investment} \\
+ \text{ Public Investment} \\
+ \text{ (Exports - Imports)} \\
+ \text{ Change in inventories} \\
+ \text{ Statistical Discrepancy} 
\end{aligned}
$$

- Note: For ease of presentation, we ignore the statistical discrepancy and create a net export term (ie, exports less imports).

## Components of GDP (E) 

```{r}
f1 <- gdp_parts %>% select.(-`GDP (E)`, -`GDP (E) - Sum of Parts`, -dif, -`Statisitical Discrepancy`, -Imports, -Exports) %>% 
  pivot_longer.(cols = c(-date)) %>% 
  ggplot(aes(x=date, y = value)) + geom_line(size = 1.5) + 
  theme_minimal() +
  theme(legend.position="bottom", legend.title=element_blank()) +
  scale_y_continuous(labels = comma) + 
  scale_color_manual(values=c("#000000")) + 
  labs(y = "$m", x = "", 
       title = "Main Components of GDP (Expenditure)", 
       caption = "Source: ABS 5204, Table 2") + 
  facet_wrap("name")
ggplotly(f1) %>%  layout(legend = list(orientation = "h", y = -0.1, x = 0.1))
```

---

```{r}

f2 <- gdp_parts %>% select.(-`GDP (E)`, -dif, -`Statisitical Discrepancy`, -Imports, -Exports) %>% 
  mutate.(ggc = 100*(`General Government Consumption` / `GDP (E) - Sum of Parts`), 
          hc = 100*(`Household Consumption` / `GDP (E) - Sum of Parts`),
          iv = 100*(Inventories / `GDP (E) - Sum of Parts`),
          nx = 100*(`Net Exports` / `GDP (E) - Sum of Parts`),
          pri = 100*(`Private Investment` / `GDP (E) - Sum of Parts`),
          pub = 100*(`Public Investment` / `GDP (E) - Sum of Parts`)) %>% 
  select.(-`General Government Consumption`, -`Household Consumption`, -Inventories, -`Net Exports`, -`Private Investment`, -`Public Investment`, -`GDP (E) - Sum of Parts`) %>% 
  rename.(`General Government Consumption` = ggc, 
          `Household Consumption` = hc, 
          Inventories = iv, 
          `Net Exports` = nx,
          `Private Investment` = pri, 
          `Public Investment` = pub) %>% 
  pivot_longer.(cols = c(-date)) %>% 
  filter.(date >= "1994-06-01") %>% 
  ggplot(aes(x=date, y = value, fill = name)) + geom_col() +
  scale_fill_manual(values=c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2")) +
  theme_minimal() +
  theme(legend.position="bottom", legend.title=element_blank()) +
  scale_y_continuous(labels = comma) + 
  labs(y = "(%)", x = "", 
       title = "Main Components of GDP (Expenditure) as a Percentage of Total", 
       caption = "Source: ABS 5204, Table 2") 
ggplotly(f2) %>%  layout(legend = list(orientation = "h", y = -0.1, x = 0.1))
```

## Summary 

- GDP (E) is the sum of household and government final consumption, public and private investment, net exports, and inventories.
- ABS publishes volume estimates of annual values in the Australian System of National Accounts (ASNA – cat no 5204)
- While there are 174 columns in the relevant table (ie, no 2), there are only 8 relevant to the total and the largest component is household final consumption expenditure. 

# Extras

## Chain volume estimates and non-additivity

- Economic analysis is about the allocation of scare resources. So if we just use current prices, we’re not shedding light on the volume of resources consumed/produced. 
- Two options: 
  - **Constant prices**: Deflate all values via price index(es) which use a common base period. Often this is expressed as using the values of a particular year, such as $USD2015.
  - **Chain volumes**: This involves estimating the quantities in each overlapping years in the same prices; that is, estimate the 2019 and 2020 values using 2019 prices or vice versa.


## Chain volume estimates and non-additivity

- Chain volume measures value quantities by prices in a base period which is updated annually. That base period for national accounts is always the period or year prior to the current period. 
  - So if you’re looking at the 2020 national accounts, the value for 2019 in chain volumes is equal to the 2019 current price value. And so on, if you’re looking at the 2030 national accounts, the value for 2029 in chain volumes is equal to the 2029 current price value.
  - The components of GDP only add up in the most recent periods. Hence the further we are away from the most recent period, the bigger the divergence between the sum of the parts and the whole. 

## Chain volume estimates and non-additivity
- In other words, updating the reference period means revisions to the *levels* of chain volume measures for all subsequent periods. 
- Hence the chain volume estimate for each year will change in each subsequent release but growth rates don’t.
- So changing the reference period doesn’t change the growth rate. And that’s the whole point. We’re prioritising growth rates over levels (of GDP). 
- For worked examples see the [ABS explanation](https://www.abs.gov.au/ausstats/abs@.nsf/0/95ce2d6796bd15aeca256db800754639/$FILE/ATT4T7WF/Demystifying%20Chain%20Volume%20Measures_1.pdf)

## Example: Non-additivity
```{r}
gdp_parts %>% select.(`GDP (E) - Sum of Parts`, `GDP (E)`, date) %>% 
   pivot_longer.(cols = c(-date)) %>% 
  ggplot(aes(x=date, y = value, group = name, colour = name)) + geom_line(size = 1.5) + 
  theme_minimal() +
  theme(legend.position="bottom", legend.title=element_blank()) +
  scale_y_continuous(labels = comma) + 
  scale_color_manual(values=c("#000000", "#E69F00")) +
  labs(y = "$m", x = "", 
      title = "Non-Additivity of the Components of GDP (Expenditure)", 
      subtitle = "Additivity refers to an aggregate being the sum of its components. This only exists in volume index estimates when \na fixed set of prices is used. As weights of a chain volume index change from year to year, chain volume indexes \nhave no base period",
      caption = "Source: ABS 5204, Table 2") + 
  annotate("text", x = as.Date("1975-06-01"), y = 1000000, size = 3, fontface = "bold",
           label = "The further away from the current reference period, \nthe more the sum of the parts diverges from the total") +
annotate("segment", x = as.Date("1970-06-01"), xend = as.Date("1970-06-01"), y = 800000, yend = 500000,
         colour = "#000000", size = 1.5, arrow = arrow())

```