---
title: "Introduction to the Australian System of National Accounts"
subtitle: "Module 2: Gross Domestic Product (Income)"
author: "David Hansell"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(data.table)
library(tidytable)
library(kableExtra)
library(readabs)
library(tidytable)
library(ggplot2)
library(tidyverse)
library(scales, lubridate)
library(plotly)

coe <- read_abs(series_id = "A2421040T") %>% 
  select.(value, date) %>% 
  rename.(coe = value)

gos <- read_abs(series_id = "A2421313L") %>% 
  select.(value, date) %>% 
  rename.(gos = value)

gmi <- read_abs(series_id = "A2421314R") %>% 
  select.(value, date) %>% 
  rename.(gmi = value)

tls <- read_abs(series_id = "A2421316V") %>% 
  select.(value, date) %>% 
  rename.(tls = value)

sd <- read_abs(series_id = "A2421018W") %>% 
  select.(value, date) %>% 
  rename.(sd = value)

ipart <- coe %>% 
  full_join.(gos) %>% 
  full_join.(gmi) %>% 
  full_join.(tls) %>% 
  rename.(`Compensation of Employees` = coe, 
          `Gross Operating Surplus` = gos, 
          `Gross Mixed Income` = gmi, 
          `Taxes Less Subsidies` = tls)
```

## Aims of the module

1. Understand the income approach of GDP 
2. How to map these concepts to ABS national accounts time series


Note: there is a corresponding module that demonstrates how to download ABS data directly in R-Studio and make the charts presented here.

## Overview

- GDP = Gross Domestic Product
  + **Gross** = includes depreciation of capital
  + **Domestic** = confined to a production boundary
  + **Product** = what is produced
- System of National Accounts (SNA) defines GDP three ways
  + **Expenditure**: Consumption + Investment + Govt Purchases + Exports – Imports 
  + **Income**: Compensation of Employees (CoE) + Gross Operating Surplus (GOS)
  + **Production**: Output less intermediate Expenses (value-added). 
- In other words, we spend our income on what we produce.

## Overview

This module focusses on the income approach or GDP(I). 

We divide the module into two parts:  

1. Concepts  
2. Measurement.
  
# Concepts

## GDP (I) - Concepts

Income approach to GDP is based on the framework: 
$$
\small
\begin{aligned}
GDP = \text{Compensation of Employees (CoE)}   \\
+ \text{ Gross Operating Surplus (GOS)} \\
+ \text{ Gross Mixed Income (GMI)} \\
+ \text{ Taxes} \\
- \text{ Subsidies (on Production and Imports)} 
\end{aligned}
$$

$$
\small
\left.
\begin{array}{l}
\text{CoE is returns to labour}\\
\text{GOS is returns to capital}\\
\text{GMI is combined labour and capital}
\end{array}
\right\}
=\text{ Income to Factors}
$$

## GDP (I) - Concepts 

- **Compensation of Employees** – includes wages and salaries paid to employees and employer social contributions.
  - Employer’s social contributions are paid by employers to employees (typically) in advance of the benefit being paid. These include payments for sickness, accidents, redundancy, retirement and superannuation. 
- **Gross Operating Surplus** is the surplus paid to owners of incorporated businesses. It includes profits, depreciation and obsolecence of capital. It’s often measured as a residual: ie, once we know CoE and the level of GDP, we can back-out GoS. 

## GDP (I) - Concepts 

- **Gross Mixed Income** is best thought of as CoE and GOS for unincorporated (and small) businesses. 
  - For example, if a business owner doesn’t pay a wage to his/herself but receives “profits” from the business, this is a combination of the return to capital and labour, but the returns to factors of production (capital and labour) are combined.
- **Total Factor Income** is the sum of Compensation of Employees, Gross Operating Surplus and Gross Mixed Income. 


## GDP (I) - Concepts 

- **Taxes less subsidies** on production and imports – this item is necessary to convert GDP at basic prices – ie returns to the factors of production – into GDP at market prices. 
  - Why? Because GDP(E) is measured in market prices, which include taxes and subsidies on the products consumed or transformed into capital.
  
# Measurement

## Measurement - Overview

- ABS publishes [annual (5204)](https://www.abs.gov.au/statistics/economy/national-accounts/australian-system-national-accounts/latest-release) and [quarterly (5206)](https://www.abs.gov.au/statistics/economy/national-accounts/australian-national-accounts-national-income-expenditure-and-product/latest-release) national accounts. 
- Quarterly estimates are often carried forward by projecting the movements in correlated series and are subject to more revision than annual estimates 
- Annual estimates (apart from the most recent year) are balanced via the supply-use tables. 
  - Main point: Once the level of GDP is finalized the statistical discrepancy item is zero. Supply-use tables start in 1994 so before this the statistical discrepancy is not zero.
- Hence here we focus on the annual estimates found in table 6 of catalogue 5204.

## Measurement - Overview

- Table 6 contains current price estimates of income. 
  - Why current prices and not a volume measure? 
    - No natural price index for GOS. 
    - Wages can be deflated separately via Wage Price Index or GDP deflator.
    - Hours worked are also measured and that is a pure volume measure.
- Compared to expenditure approach, there are not as many series in table 6. 

## Measurement - Compensation of Employees

- The two series that make up Compensation of Employees are
  - **Wages and Salaries**: Series ID = A2421038F
  - **Employer Social Contributions**: Series ID = A2421039J
- However these two series only start from the 1989-90 fiscal year. 
- Hence we use **Compensation of Employees** so we can look at longer term trends: 
  - Series ID = A2421040T

## Measurement - Gross Operating Surplus

- **Gross operating surplus** (series ID = A2421313L ) consists of 5 main series
  - Private non-financial corporations (A2421041V)
  - Public non-financial corporations (A2421042W)
  - Financial corporations (A2421309W)
  - General government (A2421311J)
  - Dwellings owned by persons (A2421312K)
- The most confusing of these is dwellings.
  - Main idea: Dwellings are capital and provide services (ie, shelter).
  
## Measurement - GMI and Taxes/Subsidies

- **Gross Mixed Income** (series ID = A2421314R) is for unincorporated enterprises where it’s difficult to separate a return to capital and return to labour. 
  - Example: A family operates a business as a trust that pays no official wages, but provides dividends to the members of the trust.
- As discussed above, **taxes less subsidies on production and imports** (series ID = A2421316V) is needed to convert GDP at basic prices to market prices. 

## Measurement - Components of GDP(I)
```{r}
f1 <- ipart %>%   pivot_longer.(cols = c(-date)) %>% 
  # filter.(date >= "2000-06-01") %>% 
  ggplot(aes(x=date, y = value)) + geom_line(size = 1.5) + 
  theme_minimal() +
  theme(legend.position="bottom", legend.title=element_blank()) +
  scale_y_continuous(labels = comma) + 
  scale_color_manual(values=c("#000000")) + 
  labs(y = "$m", x = "", 
       title = "Main Components of GDP (Income)", 
       caption = "Source: ABS 5204, Table 6") + 
  facet_wrap("name")
ggplotly(f1) %>%  layout(legend = list(orientation = "h", y = -0.1, x = 0))

```

## Measurement - Components of GDP(I)

```{r}
f2 <- ipart %>% 
  mutate.(tfi = `Compensation of Employees` + `Gross Operating Surplus` + `Gross Mixed Income`, 
          coe = 100*(`Compensation of Employees` / tfi), 
          gos = 100*(`Gross Operating Surplus` / tfi), 
          gmi = 100*(`Gross Mixed Income` / tfi)) %>% 
  select.(-`Compensation of Employees`, -`Gross Operating Surplus`, -`Gross Mixed Income`, -tfi, -`Taxes Less Subsidies`) %>% 
  rename.(`Compensation of Employees` = coe, 
          `Gross Operating Surplus` = gos, 
          `Gross Mixed Income` = gmi) %>% 
  pivot_longer.(cols = c(-date)) %>% 
  ggplot(aes(x=date, y = value, group = name, colour = name)) + geom_line(size = 1) +
  theme_minimal() +
  theme(legend.position="bottom", legend.title=element_blank()) +
  scale_colour_manual(values=c("#000000", "#E69F00", "#56B4E9")) +
  labs(y = "(%)", x = "", 
       title = "Main Components of GDP (Income) as a Percentage of Total Factor Income", 
       caption = "Source: ABS 5204, Table 6") 
ggplotly(f2) %>%  layout(legend = list(orientation = "h", y = -0.1, x = 0.1))
```


## Summary 

- The income approach to GDP shows the returns to the factors of production: labour and capital.
  - Note: taxes less subsidies on production and imports is needed to ensure consistency of GDP(I) and the other measures of GDP
- The largest share of GDP(I) is compensation of employees.
- The share of gross operating surplus has risen over time, largely due to the fall in gross mixed income. 
  - This doesn’t necessarily reflect a rise in corporate profit, but a change in business structures away from unincorporated business to incorporated entities.

# Extras

## Extras

```{r}
f3 <- ipart %>% 
  mutate.(tfi = `Compensation of Employees` + `Gross Operating Surplus` + `Gross Mixed Income`, 
          coe = 100*(`Compensation of Employees` / tfi), 
          gos = 100*(`Gross Operating Surplus` / tfi), 
          gmi = 100*(`Gross Mixed Income` / tfi)) %>% 
select.(-`Compensation of Employees`, -`Gross Operating Surplus`, -`Gross Mixed Income`, -tfi, -`Taxes Less Subsidies`) %>% 
  rename.(`Compensation of Employees` = coe, 
          `Gross Operating Surplus` = gos, 
          `Gross Mixed Income` = gmi) %>% 
  pivot_longer.(cols = c(-date)) %>% 
  ggplot(aes(x=date, y = value, fill = name)) + geom_col(position = position_stack(reverse = TRUE)) +
  scale_fill_manual(values=c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2")) +
  theme_minimal() +
  theme(legend.position="bottom", legend.title=element_blank()) +
  scale_y_continuous(labels = comma) + 
  labs(y = "(%)", x = "", 
       title = "Main Components of GDP (Income) as a Percentage of Total Factor Income", 
       caption = "Source: ABS 5204, Table 6") 
ggplotly(f3) %>%  layout(legend = list(orientation = "h", y = -0.1, x = 0.1))
```

